Efficient UQ for chemical systems using active \author{Manav Vohra, Alen Alexanderian, Hayley Guy, Sankaran Mahadevan} subspaces

### 2018-07-21 Paper/assa.tex
..

\subsection{Active subspaces}
We recall some background regarding active subspaces~\cite{Constantine15}.
\subsection{Global sensitivity measures}
Define the total Sobol' indices $T_i(f)$ and discuss their use.

For $f: \Omega \to \R$, we consider the DGSMs~\cite{SobolKucherenko09},
\[
    \nu_i(f) := \E{\left(\frac{\partial f}{\partial\xi_i}\right)^2} =
                  \int_\Omega 
                  \left(\frac{\partial f}{\partial\xi_i}\right)^2
                  \pi(\vec{\xi})d\vec{\xi}, \quad i = 1, \ldots, \Np.   
\]
Here $\pi$ is the joint PDF of $\xi$. 
Let $\mat{C} = \E{\nabla f \nabla f^T}$ and note that
\[
   \nu_i(f) = \vec{e}_i^T \mat{C} \vec{e}_i, 
\]
where $\vec{e}_i$ is the $i$th coordinate vector in $\R^\Np$.
This also shows that $\sum_{j=1}^\Np \nu_i(f) = \trace(\mat{C})$.
We consider the spectral decomposition of $\mat{C}$, given by  
$\mat{C} = \sum_{k=1}^\Np \lambda_k \vec{u}_k \vec{u}_k^T$, where 
$\lambda_i$ are the (non-negative) eigenvalues of $\mat{C}$ and
$\vec{u}_k$ are the corresponding (orthonormal) eigenvectors.
We note that,
\begin{equation}\label{equ:spectral_DGSM}
\nu_i(f) = \vec{e}_i^T \Big(\sum_{k=1}^\Np \lambda_k \vec{u}_k \vec{u}_k^T\Big) \vec{e}_i
 = \sum_{k=1}^\Np \lambda_k \ip{\vec{e}_i}{\vec{u}_k}^2, 
\end{equation}
which gives a spectral representation for the DGSMs. Truncating the summation 
leads to the notion of activity scores introduced in~\cite{Diaz16,ConstantineDiaz17}
defined as follows: 
\[
   \alpha_i(f; r) =  \sum_{k=1}^r \lambda_k \ip{\vec{e}_i}{\vec{u}_k}^2,
   \quad j = 1, \ldots, \Np, \quad r \leq \Np.
\]
The activity scores connect ideas from active subspaces and global sensitivity
analysis, and can be used to approximate DGSMs.  The following result, which
can be found in~\cite{Diaz16,ConstantineDiaz17}, quantifies the error in this
approximation. We provide a short proof for completeness. 
\begin{proposition}\label{prp:dgsm_bound} 
For $1 \leq r \leq \Np$,
\[
0 \leq \nu_i(f) - \alpha_i(f; r) \leq \lambda_{r+1}, \quad i = 1, \ldots, \Np.
\] 
\end{proposition}
\begin{proof}

@@

Consider a function $y = f(\xi_1, \xi_2, \ldots, \xi_\Np)$ where
$\xi_i$ are independent are either uniformly distributed or
distributed according to a Boltzman probability distribution.  
It was shown in~\cite{LamboniIoossPopelinEtAl13} that the total Sobol' 
index $T_i(f)$ can be bounded in terms of $\nu_i(f)$:
\begin{equation}\label{equ:sobol_bound}
T_i(f) \leq \frac{C_i}{\V(f)}\nu_i(f), \quad i = 1, \ldots, \Np,
\end{equation}
where for each $i$, $C_i$ is an appropriate \emph{Poincar\'{e}} constant.
This provides a strong theoretical basis for using DGSMs to identify 
unimportant inputs. Moreover, 
in~\cite{VohraAlexanderianSaftaEtAl18} we introduced the normalized screening metric
\[
   \tilde{\nu}_i(f) = \frac{C_i \nu_i(f)}{\sum_{i=1}^\Np C_i \nu_i(f)},
\]
for detecting unimportant input parameters. 
Note that if $\xi_i$ are iid, the $C_i$'s will cancel and the normalized screening
metric reduces to 
\[
    \tilde{\nu}_i(f) = \frac{\nu_i(f)}{\sum_{i=1}^\Np \nu_i(f)} 
      = \frac{\vec{e}_i^T \mat{C} \vec{e}_i}{\trace(\mat{C})} 
      = \frac{\sum_{k=1}^\Np \lambda_k \ip{\vec{e}_i}{\vec{u}_k}^2}{\sum_{k = 1}^\Np \lambda_k}.
\]
The following result is a straightforward generalization of
the result from~\cite{Diaz16}, and 
links activity scores and Sobol' indices.
\begin{proposition}
For $1 \leq r \leq \Np$, and $i = 1, \ldots, \Np$, we have
\[
       T_i(f) \leq \frac{C_i}{\V(f)}(\alpha_i(f; r) + \lambda_{r+1}).
\]
\end{proposition}
\begin{proof}
The result follows immediately from Proposition~\ref{prp:dgsm_bound} and~\eqref{equ:sobol_bound}.
\end{proof}
We can also bound the normalized DGSMs using activity scores; it is easy to see  
\[
\tilde{\nu}_i(f) \leq 
\frac{ C_i \big(\alpha_i(f; r) + \lambda_{r+1}\big)}{\sum_{i=1}^\Np C_i \alpha_i(f; r)}
=\frac{C_i \alpha_i(f; r)}{\sum_{i=1}^\Np C_i \alpha_i(f; r)} + \kappa_i \lambda_{r+1}, 
\]
whith $\kappa_i = C_i / (\sum_i C_i \alpha_i(f; r))$. 
In the case where where $\lambda_{r+1} \approx 0$, 
this motivates definition of
normalized activity scores
\[
   \tilde{\alpha}_i(f; r) =  \frac{C_i \alpha_i(f; r)}{\sum_{i=1}^\Np C_i \alpha_i(f; r)}.
\]

>>>>

\subsection{Active subspaces}
\label{sub:ac}
We recall some background regarding active subspaces~\cite{Constantine:2015}.

\subsection{Global sensitivity measures}
\label{sub:gsa}
Define the total Sobol' indices $T_i(f)$ and discuss their use.

For $f: \Omega \to \R$, we consider the DGSMs~\cite{Sobol:2009},
\[
    \nu_i(f) := \E{\left(\frac{\partial f}{\partial\xi_i}\right)^2} =
                  \int_\Omega 
                  \left(\frac{\partial f}{\partial\xi_i}\right)^2
                  \pi(\vec{\xi})d\vec{\xi}, \quad i = 1, \ldots, \Np.   
\]
Here $\pi$ is the joint PDF of $\xi$. 
Let $\mat{C} = \E{\nabla f \nabla f^T}$ and note that
\[
   \nu_i(f) = \vec{e}_i^T \mat{C} \vec{e}_i, 
\]
where $\vec{e}_i$ is the $i$th coordinate vector in $\R^\Np$.
This also shows that $\sum_{j=1}^\Np \nu_i(f) = \trace(\mat{C})$.
We consider the spectral decomposition of $\mat{C}$, given by  
$\mat{C} = \sum_{k=1}^\Np \lambda_k \vec{u}_k \vec{u}_k^T$, where 
$\lambda_i$ are the (non-negative) eigenvalues of $\mat{C}$ and
$\vec{u}_k$ are the corresponding (orthonormal) eigenvectors.
We note that,
\begin{equation}\label{equ:spectral_DGSM}
\nu_i(f) = \vec{e}_i^T \Big(\sum_{k=1}^\Np \lambda_k \vec{u}_k \vec{u}_k^T\Big) \vec{e}_i
 = \sum_{k=1}^\Np \lambda_k \ip{\vec{e}_i}{\vec{u}_k}^2, 
\end{equation}
which gives a spectral representation for the DGSMs. Truncating the summation 
leads to the notion of activity scores introduced in~\cite{Diaz:2016,Constantine:2017}
defined as follows: 
\[
   \alpha_i(f; r) =  \sum_{k=1}^r \lambda_k \ip{\vec{e}_i}{\vec{u}_k}^2,
   \quad j = 1, \ldots, \Np, \quad r \leq \Np.
\]
The activity scores connect ideas from active subspaces and global sensitivity
analysis, and can be used to approximate DGSMs.  The following result, which
can be found in~\cite{Diaz:2016,Constantine:2017}, quantifies the error in this
approximation. We provide a short proof for completeness. 
\begin{proposition}\label{prp:dgsm_bound} 
For $1 \leq r \leq \Np$,
\[
0 \leq \nu_i(f) - \alpha_i(f; r) \leq \lambda_{r+1}, \quad i = 1, \ldots, \Np.
\] 
\end{proposition}
\begin{proof}

@@

Consider a function $y = f(\xi_1, \xi_2, \ldots, \xi_\Np)$ where
$\xi_i$ are independent are either uniformly distributed or
distributed according to a Boltzman probability distribution.  
It was shown in~\cite{Lamboni:2013} that the total Sobol' 
index $T_i(f)$ can be bounded in terms of $\nu_i(f)$:
\begin{equation}\label{equ:sobol_bound}
T_i(f) \leq \frac{C_i}{\V(f)}\nu_i(f), \quad i = 1, \ldots, \Np,
\end{equation}
where for each $i$, $C_i$ is an appropriate \emph{Poincar\'{e}} constant.
This provides a strong theoretical basis for using DGSMs to identify 
unimportant inputs. Moreover, 
in~\cite{Vohra:2018} we introduced the normalized screening metric
\[
   \tilde{\nu}_i(f) = \frac{C_i \nu_i(f)}{\sum_{i=1}^\Np C_i \nu_i(f)},
\]
for detecting unimportant input parameters. 
Note that if $\xi_i$ are iid, the $C_i$'s will cancel and the normalized screening
metric reduces to 
\[
    \tilde{\nu}_i(f) = \frac{\nu_i(f)}{\sum_{i=1}^\Np \nu_i(f)} 
      = \frac{\vec{e}_i^T \mat{C} \vec{e}_i}{\trace(\mat{C})} 
      = \frac{\sum_{k=1}^\Np \lambda_k \ip{\vec{e}_i}{\vec{u}_k}^2}{\sum_{k = 1}^\Np \lambda_k}.
\]
The following result is a straightforward generalization of
the result from~\cite{Diaz:2016}, and 
links activity scores and Sobol' indices.
\begin{proposition}
For $1 \leq r \leq \Np$, and $i = 1, \ldots, \Np$, we have
\[
       T_i(f) \leq \frac{C_i}{\V(f)}(\alpha_i(f; r) + \lambda_{r+1}).
\]
\end{proposition}
\begin{proof}
The result follows immediately from Proposition~\ref{prp:dgsm_bound} and~\eqref{equ:sobol_bound}.
\end{proof}
We can also bound the normalized DGSMs using activity scores; it is easy to see  
\[
\tilde{\nu}_i(f) \leq 
\frac{ C_i \big(\alpha_i(f; r) + \lambda_{r+1}\big)}{\sum_{i=1}^\Np C_i \alpha_i(f; r)}
=\frac{C_i \alpha_i(f; r)}{\sum_{i=1}^\Np C_i \alpha_i(f; r)} + \kappa_i \lambda_{r+1}, 
\]
whith $\kappa_i = C_i / (\sum_i C_i \alpha_i(f; r))$. 
In the case where where $\lambda_{r+1} \approx 0$, 
this motivates definition of
normalized activity scores
\[
   \tilde{\alpha}_i(f; r) =  \frac{C_i \alpha_i(f; r)}{\sum_{i=1}^\Np C_i \alpha_i(f; r)}.
\]

### 2018-07-21 Paper/intro.tex
..

Time evolution of a chemically reacting system is largely dependent upon rate constants
associated with individual reactions. The rate constants are typically assumed to exhibit
a certain correlation with temperature (e.g. Arrhenius-type). Hence, accurate specification
of the rate-controlling parameters is critical to the fidelity of simulations. However,
in practical applications, the parameters are either specified using expert knowledge or
estimated based on a regression-fit to a set of sparse and noisy data. Consequently, the
underlying uncertainty associated with the parameters and therefore the simulation-prediction
is typically ignored. Vohra et al. calibrated the Arrhenius diffusion parameters to quantify
the uncertainty in mixing rates in Zr/Al reactive multilayers using a Bayesian framework
in~\cite{Vohra:2014} and~\cite{Vohra:2017}. The framework accounted for prior knowledge,
measurement noise, and the discrepancy between model and experiments. Bayesian calibration
was accelerated in these efforts using a polynomial chaos surrogate for the model. However,
the number of uncertain parameters for calibration were limited to two in both cases. In
higher dimensions, the construction of surrogate as well as the calibration process become
remarkably challenging.

>>>>

Time evolution of a chemically reacting system is largely dependent upon rate constants
associated with individual reactions. The rate constants are typically assumed to exhibit
a certain correlation with temperature (e.g. Arrhenius-type). Hence, accurate specification
of the rate-controlling parameters is critical to the fidelity of simulations. However,
in practical applications, the parameters are either specified using expert knowledge or
estimated based on a regression-fit to a set of sparse and noisy data. Consequently, the
underlying uncertainty associated with the parameters and therefore the simulation-prediction
is typically ignored. Vohra et al. calibrated the Arrhenius diffusion parameters to quantify
the uncertainty in mixing rates in Zr/Al reactive multilayers using a Bayesian framework
in~\cite{Vohra:2014} and~\cite{Vohra:2017}. The framework accounted for prior knowledge,
measurement noise, and the discrepancy between model and experiments. Bayesian calibration
was accelerated in these efforts using a polynomial chaos~\cite{Ghanem:2003, Xiu:2002} 
surrogate for the model. However,
the number of uncertain parameters for calibration were limited to two in both cases. In
higher dimensions, the construction of surrogate as well as the calibration process become
remarkably challenging due to increase in computational effort.

In this paper, we focus our efforts on mitigating the computational challenge posed by uncertainty
quantification (UQ) for applications involving a large number of chemical reactions. Specifically, we
accomplish this by implementing an efficient approach that aims to reduce the dimensionality
of the problem by computing the so-called active subspace~\cite{Constantine:2015}. 
The active subspace, discussed later in~\ref{sub:ac}, essentially constitutes the dominant
eigenspace of a matrix derived from the gradient information of the model output with respect
to an input. The high-dimensional application problem considered in this work is the H$_2$/O$_2$
reaction mechanism~\cite{Yetter:1991}. This mechanism has been gaining a lot of attention in the
recent years as a potential source of clean energy for locomotive applications~\cite{Das:1996}.   
The mechanism involves 19 reactions including chain reactions, dissociation/recombination reactions, and
formation and consumption of intermediate species:

\begin{table}[htbp]
\renewcommand*{\arraystretch}{0.9}
\begin{center}
\begin{tabular}{llll}
\toprule
Reaction \#     & Reaction &&\\
\bottomrule
$\mathcal{R}_1$ & H + O$_2$          & $\rightleftharpoons$ & O + OH \\
$\mathcal{R}_2$ & O + H$_2$          & $\rightleftharpoons$ & H + OH \\
$\mathcal{R}_3$ & H$_2$ + OH         & $\rightleftharpoons$ & H$_2$O + H \\
$\mathcal{R}_4$ & OH + OH            & $\rightleftharpoons$ & O + H$_2$O \\
$\mathcal{R}_5$ & H$_2$ + M          & $\rightleftharpoons$ & H + H + M \\
$\mathcal{R}_6$ & O + O + M          & $\rightleftharpoons$ & O$_2$ + M \\
$\mathcal{R}_7$ & O + H + M          & $\rightleftharpoons$ & OH + M \\
$\mathcal{R}_8$ & H + OH +M          & $\rightleftharpoons$ & H$_2$O + M \\
$\mathcal{R}_9$ & H + O$_2$ + M      & $\rightleftharpoons$ & HO$_2$ + M \\
$\mathcal{R}_{10}$ & HO$_2$ + H      & $\rightleftharpoons$ & H$_2$ + O$_2$ \\
$\mathcal{R}_{11}$ & HO$_2$ + H      & $\rightleftharpoons$ & OH + OH \\
$\mathcal{R}_{12}$ & HO$_2$ + O      & $\rightleftharpoons$ & O$_2$ + OH \\
$\mathcal{R}_{13}$ & HO$_2$ + OH     & $\rightleftharpoons$ & H$_2$O + O$_2$ \\
$\mathcal{R}_{14}$ & HO$_2$ + HO$_2$ & $\rightleftharpoons$ & H$_2$O$_2$ + O$_2$ \\
$\mathcal{R}_{15}$ & H$_2$O$_2$ + M  & $\rightleftharpoons$ & OH + OH + M \\
$\mathcal{R}_{16}$ & H$_2$O$_2$ + H  & $\rightleftharpoons$ & H$_2$O + OH \\
$\mathcal{R}_{17}$ & H$_2$O$_2$ + H  & $\rightleftharpoons$ & HO$_2$ + H$_2$ \\
$\mathcal{R}_{18}$ & H$_2$O$_2$ + O  & $\rightleftharpoons$ & OH + HO$_2$ \\
$\mathcal{R}_{19}$ & H$_2$O$_2$ + OH & $\rightleftharpoons$ & HO$_2$ + H$_2$O \\
\bottomrule
\end{tabular}
\end{center}
\caption{Reaction mechanism for H$_2$/O$_2$ from~\cite{Yetter:1991}}.
\label{tab:kinetics}
\end{table}

For each reaction in table~\ref{tab:kinetics}, the reaction rate is assumed to show an Arrhenius
correlation with temperature:
%
\be
k_i(T) = A_iT^{n_i}\exp(-E_{a,i}/RT), 
\label{eq:rate}
\ee
%
where $A_i$ is the pre-exponent, $n_i$ is the index of $T$, $E_{a,i}$ is the
activation energy corresponding to the $i^{th}$ reaction, and $R$ is the
universal gas constant.  The global reaction associated with the H$_2$/O$_2$ mechanism can
be considered as follows:
%
\be
2\text{H}_2 + \text{O}_2 \rightarrow 2\text{H}_2\text{O}.
\label{eq:global}
\ee 
%
The equivalence ration ($\phi$) is given as follows:
%
\be
\phi = \frac{(M_{\text{H}_2}/M_{\text{O}_2})_\text{obs}}{(M_{\text{H}_2}/M_{\text{O}_2})_\text{st}}.
\label{eq:phi}
\ee
%
where the numerator on the right-hand-side denotes the fuel (H$_2$)-oxidizer (O$_2$) ratio at given
condition to the same quantity under stoichiometric conditions. In this study, computations were performed
at fuel-rich conditions, $\phi$ = 2.0. Homogeneous ignition at constant pressure is simulated using the
TChem software package~\cite{Safta:2011}, and the time required for the rate of temperature increase
to exceed a given threshold, regarded as `ignition delay' is recorded.

We are interested in investigating the impact of uncertainty in the rate-controlling parameters, 
pre-exponent~($A_i$) and the activation energy~($E_{a,i}$) on ignition delay. The $A_i$'s are
considered to be uniformly distributed in the interval: $[0.9A_i^\ast, 1.1A_i^\ast]$.
The $E_{a,i}$'s for all reactions except $\mathcal{R}_6$ -- $\mathcal{R}_9$ and $\mathcal{R}_{13}$
are considered to be uncertain and uniformly distributed in the interval: $[0.99E_{a,i}^\ast, 1.01E_{a,i}^\ast]$.
$A_i^\ast$ and $E_{a,i}^\ast$ are nominal estimates corresponding to the $i^\text{th}$ reaction
as provided in~\cite{Yetter:1991}. Hence, the total number of uncertain parameters is 33 which makes the
present problem reasonably high-dimensional involving a significantly large amount of computational
effort. To reduce the effort, we explore two strategies: one involving the use of gradients (grad-based)
and the other involving local linear approximation which does require gradient computation (grad-free).
These strategies are based on algorithms 1.1 and 1.2 provided  in~\cite{Constantine:2015}. The amount of 
computational effort associated with the grad-free approach is significantly smaller than the grad-based
approach. In section~\ref{sec:method}, we implement the two strategies to a simpler problem accounting for the
uncertainty only in the pre-exponents ($A_i$'s) of the rate constants in the H$_2$/O$_2$ reaction mechanism.
The two strategies are observed to yield consistent results which encourages the use of the grad-free approach
for discovering the active subspace especially in high-dimensional problems. In section~\ref{sec:app}, we present
our findings based on the implementation of the grad-free approach to the original 33-dimensional application.

The dominant eigenvectors that constitute the active subspace correspond to the important directions in the
parameter space. In other words, the variability in the model output is predominantly observed along these
directions. Moreover, the components of these eigenvectors can be used to estimate the so-called 
activity scores~\cite{Diaz:2016,Constantine:2017} which provide an insight into the relative importance
of the uncertain parameters. In 
section~\ref{sub:gsa}, we demonstrate that the activity scores could be used to approximate the 
derivative-based global sensitivity measures (DGSMs)~\cite{Sobol:2009}, and establish a relationship between
the Sobol' total effect index, the DGSMs, and the activity scores.
